% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
\begin{document}

\title{Automatic Estimation of the Optic Nerve Sheat Diameter from Ultrasound Images}
%
\titlerunning{Automatic Opti Nerve Sheat Diameter Estimation} % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{}

%
\authorrunning{} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{}
%
\institute{Kitware Inc, Carrboro NC 27510, USA,\\
\email{}
}

\maketitle              % typeset the title of the contribution

\begin{abstract}
\end{abstract}
%
\section{Introduction}
Kitware seeks to harness the integration of machine learning algorithms and
rugged, portable ultrasound devices to develop an easy-to-use system that
enables emergency responders to quickly assess the severity of a patient's
injuries at accidents and in remote environments in which expert medical
personnel and advanced diagnostic equipment are not readily available.￼

A driving application of our point-of-care, computer-assisted ultrasound system
is in-field traumatic brain injury (TBI) assessment via the detection of
increased intracranial pressure. Delayed treatment of increased intracranial
pressure can cause temporary or permanent brain damage or even long-term coma
and death.  For example, it has been shown that acute subdural hematomas in
severe TBI patients cause significant increase in intracranial pressure, are
associated with 90\% mortality if detected and treated more than 4 hours after
injury, and yet are associated with only 30\% mortality if detected and treated
earlier.

Our system detects elevated intracranial pressure associated with TBI by
automatically measuring the diameter of the optic nerve sheath, a known
indicator of elevated intracranial pressure.  

\section{Algorithm}
In a clinical setting the optic nerve sheath diameter is measured by ocular
ultrasound (see Image). An ultrasound image is acquired and the physicians
manually measures the diameter of optic nerve sheath at a location 3mm behind
the retina. Acquiring such images and making these measurements is challenging
and time consuming task.

Our algorithm reports estimates of the optic nerve sheath diameter in near real
time as an ultrasound probe is swept across the closed eye of a patient. At a
high level, the algorithm proceeds by locating the orb of the eye through
registration of an ellipse with the largest dark circle in the image data. From
the ellipse an approximate location of the optic nerve is constructed and used
to fit two bars to the walls of the optic nerve. This high-level description
leaves out several intermediate steps such as those involving image smoothing,
morphological operations, and distance transforms, that are required to achieve
good registration results. Below are images of results of the fitting procedure
which illustrate that the algorithm runs on a wide variety of images with
differing quality.


This algorithm is now integrated into a user-friendly interface and performs
estimates at 4 frames per second. Depending on the ultrasound probe, the
algorithm can be simplified by skipping the eye estimation step to run at
around 20 frames per second.

\section{Evaluation}

To evaluate the quality of the estimates we had 13 volunteers ranging from
Kitware engineers to medical professionals annotate 23 ultrasound images.  The
image below shows a linear fit of the automatic estimated diameter from two
different parameter settings of the algorithm as compared to a medical
professional. The automatic estimates had an R2 of around 0.9 and 0.8,
respectively. The mean intra correlations among medical professionals was 0.84.
Thus, on this small sample size the automatic estimates perform on par with
medical experts. The correlation among experts matches results of previous
studies.

For additional evaluation of the performance of the algorithm we built a
phantom from gelatine in which 3D-printed “optic nerves” (plastic discs) of
known diameter can be embedded under gelatine orbs and imaged, producing
ultrasound images that look surprising like clinical ocular ultrasound images.

The goal of this evaluation is to check if an accurate estimate of the optic
nerve is possible with the proposed algorithm. In this controlled setting we
first located the optic nerve using the ultrasound. Once a good image of the
optic nerve was achieved the automatic estimation process was started and run
interactively for about 10 seconds which resulted in approximately 40 to 50
optic nerve sheath diameter estimates. The table below shows that the means of
the automatic estimates are within less than +/- 0.5mm  with a relatively tight
distribution of estimates around the ground truth diameter. A study on
intra-operator variations on the same type of phantom, reports standard
deviations of the measurements ranging from 0.25 to 0.6 mm depending on
diameter size.  

\section{Conclusion}

The results presented indicate that the algorithm performs well in a controlled
setting.  A retrospective analysis of clinical images indicate that our system
performs similar to an expert, and a phantom study using 3D printed optic
nerves of known diameter suggests that our system is accurate.

The next step is to integrate an image quality estimation algorithm, so that
high quality images of the optic nerve can be automatically identified as a
probe is swept over the eye.   This will eliminate the need for the operator to
ever view, interpret, or make measurements on an ultrasound image when using it
to assess a dilated optic nerve sheath, as indicative of increased intracranial
pressure.

Our long term goal is a system that can be used by novice operators, with
minimal ultrasound experience.   The system will automatically report when a
sufficient image has been acquired, and it will automatically determine if the
diameter of the optic nerve is greater than 5mm, thereby indicating increased
intracranial pressure associated with mild or more severe TBI.


%
% ---- Bibliography ----
%
\begin{thebibliography}{}
%
\bibitem[1980]{2clar:eke}
Clarke, F., Ekeland, I.:
Nonlinear oscillations and
boundary-value problems for Hamiltonian systems.
Arch. Rat. Mech. Anal. 78, 315--333 (1982)

\bibitem[1981]{2clar:eke:2}
Clarke, F., Ekeland, I.:
Solutions p\'{e}riodiques, du
p\'{e}riode donn\'{e}e, des \'{e}quations hamiltoniennes.
Note CRAS Paris 287, 1013--1015 (1978)

\bibitem[1982]{2mich:tar}
Michalek, R., Tarantello, G.:
Subharmonic solutions with prescribed minimal
period for nonautonomous Hamiltonian systems.
J. Diff. Eq. 72, 28--55 (1988)

\bibitem[1983]{2tar}
Tarantello, G.:
Subharmonic solutions for Hamiltonian
systems via a $\bbbz_{p}$ pseudoindex theory.
Annali di Matematica Pura (to appear)

\bibitem[1985]{2rab}
Rabinowitz, P.:
On subharmonic solutions of a Hamiltonian system.
Comm. Pure Appl. Math. 33, 609--633 (1980)

\end{thebibliography}

\end{document}
